Advection Versus Diffusion in Intraventricular Transport

### 2025-02-18 paper/main.tex
reorder: methods appear after intro

@@

% Experimental section

\section{Computational and Experimental Methods}
\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model flow and transport in zebrafish larval brain ventricles, we generated a
3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo
injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. 
From the imaging data, we constructed a computational mesh of the ventricles (\Crefpanels{fig:fig1}{d}).
The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. 
The \emph{standard} mesh consisted of 132 134 tetrahedral cells and had a
maximal (minimal) edge length of 14.9 \textmu m (4.68 \textmu m). 
At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts. 
Parts of the ventricular surfaces were marked to be lined with
motile cilia (\Crefpanels{fig:fig1}{d}, magenta markers)~\cite{Olstad2019CiliaryDevelopment}.
\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b.} Particle tracking data visualizing flow directionality in the middle ventricle.
    \textbf{c.} Flow velocity magnitudes in the middle ventricle from particle tracking data.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Streamlines simulated with the baseline flow model at
    a time 25\% into the cardiac cycle. %t=0.112613 s
    \textbf{h.} Streamlines simulated with the baseline flow model at
    a time 50\% into the cardiac cycle. %t=0.225225 s
    \textbf{i.} Streamlines simulated with the baseline flow model at
    a time 75\% into the cardiac cycle. %t=0.315315 s
    \textbf{j.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries. Time instants of panels \textbf{f, h--j} are indicated.
    }
    \label{fig:fig1}
\end{figure}

\subsection{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\Crefpanels{fig:fig1}{d}, magenta markers),
we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces
of the cilia acting on the CSF.
We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$
and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the
cilia region (see~\Cref{subsec:appendixA1}).
Both $\rr$ and the value $\tau=0.65$ mPa were chosen by 
model calibration using experimental data~\cite{Olstad2019CiliaryDevelopment}.
The data indicate speeds proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and
$2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the middle ventricle, respectively.
In the anterior ventricle, the speeds were estimated at $4.7 \pm 1.3$ \textmu m/s.
We calibrated $\tau$ heuristically by comparing these experimental velocity values
with the simulated, maximum velocity magnitude.

\subsection{Flow induced by cardiac pulsations}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Crefpanels{fig:fig1}{d}, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5$ mPa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (see~\Cref{sec:appC_particle_transport}).

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}

The Stokes equations~\Cref{eq:stokes_eqs} are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization.
We set the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM) elements for the velocity.
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements~\cite{Brezzi1985TwoProblems},
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
\fixme{The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
To ensure stability of the scheme, continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.}
Since the divergence of the velocity is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly on each
element~\cite{Boffi2008FiniteProblem}, ensuring mass conservation.

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with DOLFINx~\cite{dolfinx2023preprint},
the Python interface of the finite element software
FEniCSx, with
petsc4py~\cite{Dalcin2011ParallelPython} as linear algebra backend.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we used an iterative
linear solver based on GMRES~\cite{Saad1986GMRES:Systems} with block Jacobi
preconditioning~\cite{Jacobi1845UeberGleichungen}.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}). The simulation code is openly
available~\cite{zenodo-link-to-all-code}\fixme{Add repo link.}.

%%------PHYSICAL EXPERIMENTS------%%
\subsection{Zebrafish maintenance, genotyping and strains in photoconversion experiments}
The animal facilities for zebrafish~(\emph{danio rerio}) are approved by the
Norwegian Food Safety Authority (NFSA, Mattilsynet).
The zebrafish were maintained in accordance with the guidelines set by the
NFSA and the European Communities Council Directive.

\subsection{Generation of transgenic line to study CSF \fixme{and transport} dynamics} %\emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}}
To generate a transgenic line expressing the photoconvertible protein Dendra2 in
the brain ventricles, we fused the open reading frame (ORF)
of the neuropeptide y (npy) signal peptide
(\emph{npy}: ENSDARG00000036222, Q1LW93) to the N-terminal of
zebrafish codon optimized Dendra2 DNA sequence.
The DNA sequence was synthesized with EcoR1 enzyme sites at the 5’ and
3’ ends (GenScript Biotech) and inserted into the
pT2MUASMCS vector~\cite{Asakawa2008GeneticZebrafish}
through restriction enzyme cloning (GenScript Biotech)
to generate the 5xuas:Signal-Dendra2 plasmid DNA..

>>>>

% Experimental section
\section{Computational Methods}
\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model flow and transport in zebrafish larval brain ventricles, we generated a
3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo
injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. 
From the imaging data, we constructed a computational mesh of the ventricles (\Crefpanels{fig:fig1}{d}).
The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. 
The \emph{standard} mesh consisted of 132 134 tetrahedral cells and had a
maximal (minimal) edge length of 14.9 \textmu m (4.68 \textmu m). 
At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts. 
Parts of the ventricular surfaces were marked to be lined with
motile cilia (\Crefpanels{fig:fig1}{d}, magenta markers)~\cite{Olstad2019CiliaryDevelopment}.
\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b.} Particle tracking data visualizing flow directionality in the middle ventricle.
    \textbf{c.} Flow velocity magnitudes in the middle ventricle from particle tracking data.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Streamlines simulated with the baseline flow model at
    a time 25\% into the cardiac cycle. %t=0.112613 s
    \textbf{h.} Streamlines simulated with the baseline flow model at
    a time 50\% into the cardiac cycle. %t=0.225225 s
    \textbf{i.} Streamlines simulated with the baseline flow model at
    a time 75\% into the cardiac cycle. %t=0.315315 s
    \textbf{j.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries. Time instants of panels \textbf{f, h--j} are indicated.
    }
    \label{fig:fig1}
\end{figure}

\subsubsection{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\Crefpanels{fig:fig1}{d}, magenta markers),
we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces
of the cilia acting on the CSF.
We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$
and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the
cilia region (see~\Cref{subsec:appendixA1}).
Both $\rr$ and the value $\tau=0.65$ mPa were chosen by 
model calibration using experimental data~\cite{Olstad2019CiliaryDevelopment}.
The data indicate speeds proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and
$2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the middle ventricle, respectively.
In the anterior ventricle, the speeds were estimated at $4.7 \pm 1.3$ \textmu m/s.
We calibrated $\tau$ heuristically by comparing these experimental velocity values
with the simulated, maximum velocity magnitude.

\subsubsection{Flow induced by cardiac pulsations}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Crefpanels{fig:fig1}{d}, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5$ mPa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (see~\Cref{sec:appC_particle_transport}).

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}
The Stokes equations~\Cref{eq:stokes_eqs} are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization.
We set the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM) elements for the velocity.
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements~\cite{Brezzi1985TwoProblems},
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
\fixme{The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
To ensure stability of the scheme, continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.}
Since the divergence of the velocity is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly on each
element~\cite{Boffi2008FiniteProblem}, ensuring mass conservation.

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with
DOLFINx~\cite{dolfinx2023preprint} with petsc4py~\cite{Dalcin2011ParallelPython}
as linear algebra backend.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we used an iterative
linear solver based on GMRES~\cite{Saad1986GMRES:Systems} with block Jacobi
preconditioning~\cite{Jacobi1845UeberGleichungen}. We discretize the time-dependent 
diffusion equation (without advection) and use the assembled linear system matrix
as the preconditioner matrix for the advection-diffusion problem.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}). The simulation code is openly
available~\cite{zenodo-link-to-all-code}\fixme{Add repo link.}.

%%------PHYSICAL EXPERIMENTS------%%
\section{Experimental Methods}
\subsection{Zebrafish maintenance, genotyping and strains in photoconversion experiments}
The animal facilities for zebrafish~(\emph{danio rerio}) are approved by the
Norwegian Food Safety Authority (NFSA, Mattilsynet).
The zebrafish were maintained in accordance with the guidelines set by the
NFSA and the European Communities Council Directive.

\subsection{Generation of transgenic line to study CSF \fixme{and transport} dynamics} %\emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}}
To generate a transgenic line expressing the photoconvertible protein Dendra2 in
the brain ventricles, we fused the open reading frame (ORF)
of the neuropeptide y (npy) signal peptide
(\emph{npy}: ENSDARG00000036222, Q1LW93) to the N-terminal of
zebrafish codon optimized Dendra2 DNA sequence.
The DNA sequence was synthesized with EcoR1 enzyme sites at the 5’ and
3’ ends (GenScript Biotech) and inserted into the
pT2MUASMCS vector~\cite{Asakawa2008GeneticZebrafish}
through restriction enzyme cloning (GenScript Biotech)
to generate the 5xuas:Signal-Dendra2 plasmid DNA.

@@
